---
title: "Getting Started with yolor"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with yolor}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = FALSE
)
```

## Overview

`yolor` is the training and inference companion to the
[ShinyLabel](https://github.com/Lalitgis/ShinyLabel) image annotation tool.
Together they form a fully R-native YOLO pipeline:

```
ShinyLabel  →  annotate images, save to SQLite
yolor       →  read DB, export dataset, train, detect, evaluate
```

---

## 1. Install & Setup

```{r install}
# Install yolor from GitHub
devtools::install_github("Lalitgis/yolor")

# Install the Python backend (Ultralytics + PyTorch)
library(yolor)
yolo_setup()   # creates a 'yolor' virtualenv and installs ultralytics
```

After `yolo_setup()`, restart R and activate the environment:

```{r activate}
reticulate::use_virtualenv("yolor")
library(yolor)
```

---

## 2. Annotate Images with ShinyLabel

```{r annotate}
library(shinylabel)

run_shinylabel(db_path = "my_project.db")
# → Opens the annotation UI in your browser
# → Draw bounding boxes, assign classes, mark images "done"
```

---

## 3. Read the ShinyLabel Database

```{r read}
ds <- sl_read_db("my_project.db")
print(ds)

# Check class distribution
sl_class_summary(ds)

# Plot annotation counts per class
plot(ds)
```

---

## 4. Export to YOLO Dataset

```{r export}
sl_export_dataset(
  ds,
  output_dir = "yolo_dataset/",
  val_split  = 0.2,    # 20% validation
  seed       = 42
)

# Validate the exported dataset
yolo_validate_dataset("yolo_dataset/")
```

This creates:

```
yolo_dataset/
├── data.yaml
├── images/
│   ├── train/
│   └── val/
└── labels/
    ├── train/
    └── val/
```

---

## 5. Load a YOLO Model

```{r model}
# See available pre-trained sizes
yolo_available_models()

# Load nano model (auto-downloads weights on first use)
model <- yolo_model("yolov8n")

# Or load your own weights
model <- yolo_model("path/to/weights.pt")
```

---

## 6. Train

```{r train}
result <- yolo_train(
  model,
  data    = "yolo_dataset/data.yaml",
  epochs  = 100,
  imgsz   = 640,
  batch   = 16
)

print(result)
# Best weights are at: result$best_weights
```

Training results (loss curves, confusion matrix, etc.) are saved in `runs/detect/train/`.

---

## 7. Detect Objects in New Images

```{r detect}
# Load best trained model
trained_model <- yolo_model(result$best_weights)

# Run on a directory
detections <- yolo_detect(
  trained_model,
  images = "new_photos/",
  conf   = 0.4
)

print(detections)

# Visualise on a specific image
plot(detections, image = "new_photos/img001.jpg")

# Export all detections to CSV
yolo_export_csv(detections, "detections.csv")

# Export to GeoJSON (for GIS / satellite imagery workflows)
yolo_export_geojson(detections, "detections.geojson")
```

---

## 8. Evaluate Model Performance

```{r eval}
bench <- yolo_benchmark(
  trained_model,
  data = "yolo_dataset/data.yaml"
)

print(bench)
plot(bench)   # per-class mAP bar chart
```

---

## Tips

| Situation | Recommendation |
|-----------|----------------|
| < 100 images | Use `"yolov8n"`, `epochs = 50`, `batch = 8` |
| 100–1000 images | `"yolov8s"` or `"yolov8m"`, `epochs = 100` |
| GPU available | Set `device = "cuda"` in `yolo_model()` and `yolo_train()` |
| Apple Silicon | Set `device = "mps"` |
| Low mAP | Increase `epochs`, check class balance with `sl_class_summary()` |
| Slow training | Reduce `imgsz` to `320` or set `cache = TRUE` |

---

## Full Minimal Example

```{r full}
library(yolor)
reticulate::use_virtualenv("yolor")

# Read ShinyLabel DB
ds <- sl_read_db("project.db")

# Export YOLO dataset
sl_export_dataset(ds, "dataset/", val_split = 0.2)

# Train
model  <- yolo_model("yolov8n")
result <- yolo_train(model, data = "dataset/data.yaml", epochs = 50)

# Detect
preds <- yolo_detect(result, images = "test_images/", conf = 0.3)
plot(preds)
```
